Resonant Scattering in a Uniform Sphere with Large Optical Depth

### 2021-09-22 ms.tex
Residue theorem section made more clear / explicit

The source is changed to the form
\be
j_{\rm em} & = & \frac{E}{4\pi} \delta^3(\vec{x} - \vec{x}_s) \delta(\nu-\nu_{\rm s})\delta (t) ,
\label{eq:jem2}
\ee
which is a standard expression for emissivity in CGS units where $E$ is the energy emitted in the impulse at time $t=0$. 
The resulting equation for $J(r,\sigma,t)$ is
\be
-3 \frac{k\phi}{c} \frac{\partial J}{\partial t} + \nabla^2 J + \left( \frac{k}{\Delta} \right)^2 \frac{\partial^2 J}{\partial \sigma^2}
& = & - \frac{\sqrt{6} kE}{4\pi \Delta^2} \delta^3(\vec{x}) \delta (\sigma - \sigma_s ) \delta (t).
\label{eq:diffusion_eqn}
\ee
We employ an expansion in terms of spherical Bessel functions in $r$ and transform the time dependence into an imaginary frequency variable $\omega$. The eigenvalues $i\omega$ are real and positive for damped solutions to the time-dependent equation. The expansion for $J(r, \sigma, t)$ is
\be
\label{eq:jrsigmat_expansion}
J(r, \sigma, t) = -\sum_{n=1}^{\infty} \int \frac{d\omega}{2\pi} e^{-i\omega t} j_0\left(\kappa_n r\right) J(n, \sigma, \omega),
\ee
with
\be \label{eq:jnsigmaomega}
J(n, \sigma, \omega) = \frac{2\kappa_n^2}{R} \int_0^R dr\ r^2 j_0(\kappa_n r) \int_0^\infty dt\ e^{i\omega t} J(r, \sigma, t).
\ee
Eq. \ref{eq:jrsigmat_expansion} is the standard specific mean intensity $dE/dA dt d\nu$ in CGS units. %energy per unit area per unit time per Hertz, expressed here in CGS units.
Though $J(r, \sigma, t)$ is written as a function of the photon frequency variable $\sigma$, it should be integrated over $d\nu$ to preserve its normalization. Eq. \ref{eq:jnsigmaomega} is in the same units, but with an additional unit of $t$ in the numerator as a result of the expansion. Plugging in Eq. \ref{eq:jnsigmaomega} for the eigenfunctions in Eq. \ref{eq:diffusion_eqn} and evaluating the integrals, we obtain
\be \label{eq:diffusion_plugged_in}
 \left( \frac{3k\phi}{c}i\omega  -   \kappa_n^2 \right) J(n,\sigma,\omega)  &+& \left( \frac{k}{\Delta} \right)^2 \frac{\partial^2J(n,\sigma,\omega)}{\partial\sigma^2} = -\frac{2\kappa_n^2}{R} \frac{\sqrt{6}}{4\pi} \frac{kE}{\Delta^2} \frac{1}{4\pi} \delta(\sigma - \sigma_s).
\ee
This cannot be solved analytically, so we turn now to a semi-analytic scheme to solve this equation for $J$. We start with a treatment of the boundary conditions. At $\sigma=\sigma_s$ continuity must be enforced,
\be \label{eq:matching_condition_1}
J_+ = J_-,
\ee
and the discontinuity in $dJ/d\sigma$  must be
\be \label{matching_condition_2}
\frac{\partial J_+}{\partial \sigma} - \frac{\partial J_-}{\partial \sigma} & = & 
- \frac{\sqrt{6}}{8} n^2 \frac{E}{kR^3}.
\ee
At large values of $\sigma$ the line profile is small such that Eq. \ref{eq:diffusion_plugged_in} becomes
\be \label{eq:diffusion_at_large_sigma}
\frac{\partial^2J}{\partial\sigma^2} = \frac{\Delta^2\kappa_n^2}{k^2} J,
\ee
which has solutions 
\be
J(n, \sigma, \omega)\ {\sim}\ e^{\pm \Delta \kappa_n \sigma / k}.
\ee
We then have
\be \label{eq:single_j_derivative}
\frac{\partial J}{\partial \sigma} = \mp \frac{\Delta\kappa_n}{k} J,
\ee
where a negative sign is taken for large $+\sigma$ and a positive sign is taken for large $-\sigma$. Two numerical integrations are performed inward from large $\pm \sigma$ to $\sigma_s$, isolating the solution that increases exponentially toward the line center. Initial values for integration are obtained by setting $J=1$ in Eq. \ref{eq:single_j_derivative}. This gives $J$ and $J'$ on either side of $\sigma_s$, where a prime indicates the derivative $\partial/\partial \sigma$. By enforcing the matching conditions, Eqs. \ref{eq:matching_condition_1} and \ref{matching_condition_2}, the solution for $J(n, \sigma, \omega)$ is obtained for all photon frequencies $\sigma$. Since the solution is linear in the starting conditions, only two integrations with different starting values are necessary.

Let us define the damping frequency to be $\gamma \equiv i\omega$, which is real-valued and positive for damped solutions. There are certain values of $\gamma$ for which a resonant response is produced which we will call $\gamma_{nm}$. Here, $n$ and $m$ are indices that represent individual spatial and frequency modes, respectively. Near these $\gamma_{nm}$, the solution will have the form 
\be
J(n,\sigma,-i\gamma) & \simeq \frac{ J_{nm}(\sigma) }{\gamma_{nm} - \gamma} + {\rm terms\ which\ vary\ slowly\ in\ \gamma}.
\ee
The contribution of the $J_{nm}(\sigma)$ eigenmode to the total solution is obtained by the residue theorem:
\be
J(r,\sigma,t) & = &  -\sum_{n=1}^{\infty}j_0(\kappa_n r)  \int \frac{d\gamma}{2\pi i} e^{-\gamma t}J(n,\sigma,-i\gamma)
\nonumber \\ & \simeq & \sum_{n=1}^{\infty}j_0(\kappa_n r)  \int \frac{d\gamma}{2\pi i} e^{-\gamma t} \frac{ J_{nm}(\sigma) }{\gamma - \gamma_{nm}} 
\nonumber \\ & \simeq & \sum_{n=1}^{\infty}j_0(\kappa_n r)  \sum_{m=1}^{\infty}J_{nm}(\sigma) e^{-\gamma_{nm}t}.
\ee
Here we have closed the integral along the imaginary $\gamma$-axis  with a semi-circle at infinity. The contribution from the semi-circle at infinity is zero, so we then obtain the residue of the $(n, m)$ pole. Summing over all spatial eigenmodes $n$ and over all temporal eigenmodes $m$ for a given $n$, we get the final form of the result
\be \label{eq:Jrsigmat}
J(r,\sigma,t) & = & \sum_{n=1}^\infty j_0(\kappa_n r)  \sum_m J_{nm}(\sigma) e^{-\gamma_{nm}t}.
\ee
Beginning with Eq. \ref{eq:Jrsigmat}, we use the derivative
\be
\frac{dj_0(\kappa_n r)}{dr} \bigg\rvert_R & =& \frac{d}{dr} \left( \frac{\sin(\kappa_n r)}{\kappa_n r} \right)\bigg\rvert_R
=  \left( \frac{\cos(\kappa_n R)}{R} - \frac{\sin(\kappa_n R)}{\kappa_n R^2} \right)
\nonumber \\ & = & \frac{(-1)^n}{R},
\ee
to obtain the flux at the surface, which is
\be
F(R,\sigma,t) & =& - \frac{4\pi}{3k\phi} \frac{dJ(R,\sigma,t)}{dr} 
= - \frac{4\pi}{3k\phi R}  \sum_{nm} (-1)^n J_{nm}(\sigma) e^{-\gamma_{nm}t}.
\ee
Multiplying by $4\pi R^2$ gives the energy per time per frequency
\be
\frac{dE}{dtd\nu} & = & - \frac{16\pi^2 R}{3k\phi}  \sum_{nm} (-1)^n J_{nm}(\sigma) e^{-\gamma_{nm}t}.
\label{eq:dEdtdnu}
\ee
Integrating this over time yields a factor $-1/\gamma_{nm}$, and over $d\nu = \sqrt{3/2} \Delta^2 \phi d\sigma$, a ``sum rule" results
\be
1 & = &  \sqrt{ \frac{3}{2} } \frac{16\pi^2R\Delta^2}{3kE} \sum_{nm} (-1)^n \gamma_{nm}^{-1} \int d\sigma J_{nm}(\sigma).
\ee
This non-trivial expression provides a check on the values of $\gamma_{nm}$ and the functions $J_{nm}(\sigma)$. This expression can also be written as
\be
1 & =& \sum_{nm} P_{nm},
\label{eq:sumrule}
\ee
where the contribution of each eigenfunction is
\be \label{eq:pnmsoln}
P_{nm} & \equiv & \sqrt{ \frac{3}{2} } \frac{16\pi^2R\Delta^2}{3kE}  (-1)^n \gamma_{nm}^{-1} \int d\sigma J_{nm}(\sigma).
\ee
These coefficients are negative for odd values of $n$ and positive for even $n$. The size of each contribution scales roughly as $0.5/(m+1/8)^{2/3}$, with a very weak dependence on $n$. This indicates the need for a large number of $n$ and $m$ to converge, i.e., it takes roughly ten times as many $m$ modes to reduce the size of $P_{nm}$ by a factor of ${\sim}$5, so convergence is slow.

>>>>

The source is changed to the form
\be
j_{\rm em} & = & \frac{E}{4\pi} \delta^3(\vec{x} - \vec{x}_s) \delta(\nu-\nu_{\rm s})\delta (t) ,
\label{eq:jem2}
\ee
which is a standard expression for emissivity in CGS units where $E$ is the energy emitted in the impulse at time $t=0$. 
The resulting equation for $J(r,\sigma,t)$ is
\be
-3 \frac{k\phi}{c} \frac{\partial J}{\partial t} + \nabla^2 J + \left( \frac{k}{\Delta} \right)^2 \frac{\partial^2 J}{\partial \sigma^2}
& = & - \frac{\sqrt{6} kE}{4\pi \Delta^2} \delta^3(\vec{x}) \delta (\sigma - \sigma_s ) \delta (t).
\label{eq:diffusion_eqn}
\ee
We employ an expansion in terms of spherical Bessel functions in $r$ and transform the time dependence into an imaginary frequency variable $\omega$. The expansion for $J(r, \sigma, t)$ is
\be
\label{eq:jrsigmat_expansion}
J(r, \sigma, t) = -\sum_{n=1}^{\infty} \int \frac{d\omega}{2\pi} e^{-i\omega t} j_0\left(\kappa_n r\right) J(n, \sigma, \omega),
\ee
with
\be \label{eq:jnsigmaomega}
J(n, \sigma, \omega) = \frac{2\kappa_n^2}{R} \int_0^R dr\ r^2 j_0(\kappa_n r) \int_0^\infty dt\ e^{i\omega t} J(r, \sigma, t).
\ee
Eq. \ref{eq:jrsigmat_expansion} is the standard specific mean intensity $dE/dA dt d\nu$ in CGS units. %energy per unit area per unit time per Hertz, expressed here in CGS units.
Though $J(r, \sigma, t)$ is written as a function of the photon frequency variable $\sigma$, it should be integrated over $d\nu$ to preserve its normalization. Eq. \ref{eq:jnsigmaomega} is in the same units, but with an additional unit of $t$ in the numerator as a result of the expansion. Plugging in Eq. \ref{eq:jnsigmaomega} for the eigenfunctions in Eq. \ref{eq:diffusion_eqn} and evaluating the integrals, we obtain
\be \label{eq:diffusion_plugged_in}
 \left( \frac{3k\phi}{c}i\omega  -   \kappa_n^2 \right) J(n,\sigma,\omega)  &+& \left( \frac{k}{\Delta} \right)^2 \frac{\partial^2J(n,\sigma,\omega)}{\partial\sigma^2} = -\frac{2\kappa_n^2}{R} \frac{\sqrt{6}}{4\pi} \frac{kE}{\Delta^2} \frac{1}{4\pi} \delta(\sigma - \sigma_s).
\ee
This cannot be solved analytically, so we turn now to a semi-analytic scheme to solve this equation for $J$. We start with a treatment of the boundary conditions. At $\sigma=\sigma_s$ continuity must be enforced,
\be \label{eq:matching_condition_1}
J_+ = J_-,
\ee
and the discontinuity in $dJ/d\sigma$  must be
\be \label{matching_condition_2}
\frac{\partial J_+}{\partial \sigma} - \frac{\partial J_-}{\partial \sigma} & = & 
- \frac{\sqrt{6}}{8} n^2 \frac{E}{kR^3}.
\ee
At large values of $\sigma$ the line profile is small such that Eq. \ref{eq:diffusion_plugged_in} becomes
\be \label{eq:diffusion_at_large_sigma}
\frac{\partial^2J}{\partial\sigma^2} = \frac{\Delta^2\kappa_n^2}{k^2} J,
\ee
which has solutions 
\be
J(n, \sigma, \omega)\ {\sim}\ e^{\pm \Delta \kappa_n \sigma / k}.
\ee
We then have
\be \label{eq:single_j_derivative}
\frac{\partial J}{\partial \sigma} = \mp \frac{\Delta\kappa_n}{k} J,
\ee
where a negative sign is taken for large $+\sigma$ and a positive sign is taken for large $-\sigma$. Two numerical integrations are performed inward from large $\pm \sigma$ to $\sigma_s$, isolating the solution that increases exponentially toward the line center. Initial values for integration are obtained by setting $J=1$ in Eq. \ref{eq:single_j_derivative}. This gives $J$ and $J'$ on either side of $\sigma_s$, where a prime indicates the derivative $\partial/\partial \sigma$. By enforcing the matching conditions, Eqs. \ref{eq:matching_condition_1} and \ref{matching_condition_2}, the eigenfunctions $J(n, \sigma, \omega)$ are obtained for all photon frequencies $\sigma$. Since the eigenfunctions are linear in the starting conditions, only two integrations with different starting values are necessary.

Let us define the damping frequency to be $\gamma \equiv i\omega$, which is real-valued and positive for damped solutions to the time-dependent equation. There are certain values of $\gamma$ for which a resonant response is produced which we will call $\gamma_{nm}$. Near these $\gamma_{nm}$ poles, the eigenfunctions will have the form 
\be \label{eq:jnsigmaomega_approx}
J(n,\sigma,-i\gamma) & \simeq \frac{ J_{nm}(\sigma) }{\gamma_{nm} - \gamma} + {\rm terms\ which\ vary\ slowly\ in\ \gamma}.
\ee
From Eq. \ref{eq:jrsigmat_expansion}, the contribution of a single pole to the total solution is
\be
J_{nm}(r,\sigma,t) & = &  -j_0(\kappa_n r)  \int \frac{d\gamma}{2\pi i} e^{-\gamma t}J(n,\sigma,-i\gamma)
\\ & \simeq & j_0(\kappa_n r)  \int \frac{d\gamma}{2\pi i} e^{-\gamma t} \frac{ J_{nm}(\sigma) }{\gamma - \gamma_{nm}}.
\ee
We close the integral along the imaginary $\gamma$-axis  with a semi-circle at infinity and apply the residue theorem. The contribution from the semi-circle is zero, so we obtain
\be
J_{nm}(r,\sigma,t) & \simeq & j_0(\kappa_n r) J_{nm}(\sigma) e^{-\gamma_{nm}t}.
\ee
Summing over all spatial modes $n$ and over all eigenmodes $m$ for a given $n$, we get the final form of the result:
\be \label{eq:Jrsigmat}
J(r,\sigma,t) & = & \sum_{n=1}^\infty j_0(\kappa_n r)  \sum_m^{\infty} J_{nm}(\sigma) e^{-\gamma_{nm}t}.
\ee
Beginning with Eq. \ref{eq:Jrsigmat}, we use the derivative
\be
\frac{dj_0(\kappa_n r)}{dr} \bigg\rvert_R & =& \frac{d}{dr} \left( \frac{\sin(\kappa_n r)}{\kappa_n r} \right)\bigg\rvert_R
=  \left( \frac{\cos(\kappa_n R)}{R} - \frac{\sin(\kappa_n R)}{\kappa_n R^2} \right)
\nonumber \\ & = & \frac{(-1)^n}{R},
\ee
to obtain the flux at the surface, which is
\be
F(R,\sigma,t) & =& - \frac{4\pi}{3k\phi} \frac{dJ(R,\sigma,t)}{dr} 
= - \frac{4\pi}{3k\phi R}  \sum_{nm} (-1)^n J_{nm}(\sigma) e^{-\gamma_{nm}t}.
\ee
Multiplying by $4\pi R^2$ gives the energy per time per frequency
\be
\frac{dE}{dtd\nu} & = & - \frac{16\pi^2 R}{3k\phi}  \sum_{nm} (-1)^n J_{nm}(\sigma) e^{-\gamma_{nm}t}.
\label{eq:dEdtdnu}
\ee
Integrating this over time yields a factor $-1/\gamma_{nm}$, and over $d\nu = \sqrt{3/2} \Delta^2 \phi d\sigma$, a ``sum rule" results
\be
1 & = &  \sqrt{ \frac{3}{2} } \frac{16\pi^2R\Delta^2}{3kE} \sum_{nm} (-1)^n \gamma_{nm}^{-1} \int d\sigma J_{nm}(\sigma).
\ee
This non-trivial expression provides a check on the values of $\gamma_{nm}$ and the functions $J_{nm}(\sigma)$. This expression can also be written as
\be
1 & =& \sum_{nm} P_{nm},
\label{eq:sumrule}
\ee
where the contribution of each eigenfunction is
\be \label{eq:pnmsoln}
P_{nm} & \equiv & \sqrt{ \frac{3}{2} } \frac{16\pi^2R\Delta^2}{3kE}  (-1)^n \gamma_{nm}^{-1} \int d\sigma J_{nm}(\sigma).
\ee
These coefficients are negative for odd values of $n$ and positive for even $n$. The size of each contribution scales roughly as $0.5/(m+1/8)^{2/3}$, with a very weak dependence on $n$. This indicates the need for a large number of $n$ and $m$ to converge, i.e., it takes roughly ten times as many $m$ modes to reduce the size of $P_{nm}$ by a factor of ${\sim}$5, so convergence is slow.
